(Deep) Induction for GADTs\vspace*{-0.1in}

### 2021-04-20 induction-paper/ENTCS-GADT-induction.tex
finished MFPS  paper draft; still to be proofread carefully

{\bf Related Work\/} Various techniques for deriving induction rules
for data types that go beyond ADTs have been studied. For example, Fu
and Selinger~\cite{fs18} show, via examples, how to derive induction
rules for arbitrary nested types. Unfortunately, however, their
technique is rather {\em ad hoc}, so is unclear how to generalize it
to nested types other than the specific ones in the
examples. Moreover, it actually derives induction rules for data types
{\em related} to the original nested types rather than for the
original nested types themselves, and it is unclear whether or not the
derived rules are sufficiently expressive to prove all results about
the original nested types that we would expect to be provable by
induction. This latter point echoes the issue with Coq-derived
induction rule for rose trees raised in Section~\ref{sec:intro}, which
has the unfortunate effect of forcing users to manually write
induction (and other) rules for such types for use in that
system. Tassi~\cite{tas19} has done exactly that, deriving induction
rules for data type definitions in Coq using unary
parametricity. Tassi's technique seems to be essentially equivalent to
that of~\cite{jp19} for nested types, although he does not permit true
nesting. More recently, Ulrich~\cite{ull20} has implemented a plugin
in MetaCoq to generate induction rules for nested types. This plugin
is also based on unary parametricity and, again, true nesting is not
permitted.  As far as we know, no attempt has yet been made to extend
either implementation to GADTs. In fact, we know of no work other than
that reported here that specifically addresses induction rules for
(deep) GADTs.

@@

Although a truly nested type admits only a single induction rule, it
is worth noting that for those nested types that do admit distinct
structural induction and deep induction rules, the latter generalizes
the former. Indeed, the structural induction rule for such a nested
type is recoverable from its deep induction rule by taking the custom
predicates on its data of primitive types to be constantly
$\mathsf{True}$-valued predicates. This instantiation ensures that the
resulting induction rule only inspects the top-level structure of its
argument, rather than the contents of that structure, which exactly
coindices with what structural induction should do.
%A concrete example of such a derivation is given in
%Section~\ref{sec:ind-equal}.

@@

The equality constraints engendered by GADTs' data constructors makes
deriving (deep) induction rules for then more involved than for ADTs
and other nested types. Nevertheless, we show in this section how to
do so. We first illustrate the key components of our approach by
deriving deep induction rules for the three specific GADTs introduced
in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
abstract these to a general framework that can be applied to any deep
GADT that is not truly nested. As hinted above, the predicate lifting
for the equality GADT plays a central role in deriving both s
tructural and deep induction rules for more general GADTs.

@@

To prove that this rule is sound we provide a witness
$\mathsf{dIndSeq}$ inhabiting the type in~\eqref{eq:ind-seq} as
follows:
\[\begin{array}{lll}
\mathsf{dIndSeq\;P\;cconst\;cpair\;A\;Q_A\;(const\,a)\;liftA}  &
    = &  \mathsf{cconst\;A\;Q_A\;a\;liftA}\\
\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,
    (Q_B,Q_C, liftE, liftB, liftC)}  & = &
\mathsf{cpair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}
\end{array}\]
In the first clause ahove, $\mathsf{a : A}$, $\mathsf{Q_A : A \to
  Set}$, $\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) =
  Q_A\,a}$. In the second, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C :
  C \to Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, $\mathsf{liftE :
  Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e}$,
$\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$, and $\mathsf{liftC :}$
$\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together ensure that
$\mathsf{(Q_B, Q_C, liftE, liftB, liftC) :
  Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$ --- and
$\mathsf{p_B} =\mathsf{dIndSeq\,P\,cconst\,cpair\,B\,Q_B\,s_B\,liftB :
  P\,B\,Q_B\,s_B}$ and $\mathsf{p_C}
=\mathsf{dIndSeq\,P\,cconst\,cpair\,C\,Q_C\,s_C\,liftC :
  P\,C\,Q_C\,s_C}$.

@@

In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
derived deep induction rules for GADTs that are not truly
nested. Since both nested types and GADTs without true nesting admit
deep induction rules, we might expect that truly nested GADTs would as
well. Surprisingly, however, they do not. That is, our results from
the previous section are the strongest possible. Indeed, the induction
rule for a data type generally relies on (unary) parametricity of the
model interpreting it, and deel induction for a truly nested type or a
ruly nested GADT also relies on this interpretation being functorial.
But, whereas nested types can admit a functorial parametric semantics,
GADTs cannot admit both functorial and parametric semantics at the
same time~\cite{jgj21}. In this section we show how deep induction for
truly nested GADTs nesting goes wrong by analyzing the following
concrete example of such a type.\looseness=-1

@@

In this section we use deep induction for the $\mathsf{LTerm}$ GADT
from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  For
this we use the following predicate, which takes a lambda term and
produces the type of its possible types:
\begin{align*}
  &\mathsf{GetType : \forall \, (A : Set) \to LTerm\,A \to Set} \\
  &\mathsf{GetType \,A \,t = Maybe \, (LType \, A)}
\end{align*}
This predicate uses $\mathsf{Maybe}$ to represent potential failure,
where $\mathsf{Maybe}$ is defined as follows:
\begin{equation}\label{eq:maybe}
\begin{array}{l}
\mathsf{data\ Maybe : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;nothing :\, Maybe\,A}\\
\mathsf{\;\;\;\;\;\;\;\;\;\;\;\;just :\, A \to Maybe\,A}
\end{array}
\end{equation}
We want to show that $\mathsf{GetType}$ is satisfied by every element
of $\mathsf{LTerm\,A}$, i.e., we want to prove:
\[ \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to
   GetType \,A \,t}\] This cannot be achieved without deep induction,
which is required to to apply the induction hyppothesis to the
individual terms in the list of terms that is an argument to the data
constructor $\mathsf{list}$ for $\mathsf{LTerm}$.

Using the deep induction rule for $\mathsf{LTerm}$ from
Section~\ref{sec:ind-lam} we define $\mathsf{getTypeProof}$ by
\[ \mathsf{getTypeProof \,A \,t = 
    dIndLTerm\, P \, cvar\, cabs\, capp\, clistc\, A\, K_\top\, t\,
    (LTerm\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
$\mathsf{P}$ is the polymorphic predicate $\mathsf{\lambda \, (A:
  Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
  A)}$ and $\mathsf{LTerm^\wedge KT : \forall\, (A : Set)\, (t : LTerm
  A) \to LTerm^{\wedge}\, A\, K_\top\,t}$ is a function that we will
define later.  We also need the applications to $\mathsf{P}$ of the
induction hypotheses from Section~\ref{sec:ind-lam} for
$\mathsf{LTerm}$'s data constructors. These are:
\[\begin{array}{lll}
\mathsf{cvar} & : & \mathsf{\forall\, (A : Set)\, (Q_A : A \to Set)\,
  (s : String)\, (T_A : LType\, A) \to LType^{\wedge} \, A\, Q_A\,
  T_A\, \to Maybe\, (LType\, A)}\\
 \mathsf{cabs} & : & \mathsf{\forall\, (A \, B\, C: Set)\, (Q_A : A
   \to Set)\, (Q_B : B \to Set)\, (Q_C : C \to Set)} \\ 
& &    \quad\mathsf{(e : Equal\, A\, (B \to C))\, (s : String)\, (T_B
   : LType\, B)\, (t_C : LTerm \, C)} \\  
& & \quad\quad\mathsf{\to Equal^{\wedge}\, A\, (B \to C)\, Q_A\,
   (Arr^{\wedge}\, B\, C\, Q_B\, Q_C)\, e  \to LType^{\wedge} \, B\,
   Q_B\, T_B  \to Maybe\, (LType\,C) \to Maybe\, (LType\, A)}\\
\mathsf{capp} & : & \mathsf{\forall\, (A \, B : Set)\, (Q_A : A \to
  Set)\, (Q_B : B \to Set)\,  (t_{BA}: LTerm\, (B \to A))\, (t_B :
  LTerm\, B)} \\  
& & \quad\mathsf{\to Maybe\, (LType\, (B \to A))\to Maybe\, (LType\,
      B) \to Maybe\, (LType\, A)}\\
\mathsf{clist} & : & \mathsf{\forall\, (A \, B : Set)\, (Q_A : A \to
  Set)\, (Q_B : B \to Set)\, (e : Equal\, A\, (List\, B))\, (ts : 
  List\, (LTerm\, B))}\\  
& & \quad\mathsf{\to Equal^{\wedge}\, A\, (List\,B) \, Q_A\,
  (List^{\wedge}\, B\, Q_B)\, e \to List^{\wedge}\, (LTerm\, B)\,
  (GetType\, B)\, ts \to Maybe\, (LType\, A)}
\end{array}\]

{\color{red} HERE!!}

For variables we simply return the type $\mathsf{T : LType\, A}$ (wrapped in $\mathsf{just}$), 
and the cases for abstraction and application are similar.
The interesting case is $\mathsf{clist}$, in which we have to use 
the results of $\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$
in order to extract the type of one of the terms in the list. 
To define $\mathsf{clist}$ we pattern-match on the list of terms $\mathsf{ts}$.
If $\mathsf{ts}$ is the empty list $\mathsf{nil}$, we cannot extract a type, 
so we return $\mathsf{nothing}$:
\[
  \mathsf{clist\, A\, B\, Q\, Q'\, e\, nil \, liftE\, lift_{ts} = nothing}
\]
where $\mathsf{liftE : Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\, e}$
and $\mathsf{lift_{ts} : List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.
If $\mathsf{ts}$ is a non-empty list $\mathsf{cons\,t\,ts'}$,
we pattern match on $\mathsf{list_{ts}}$ and use the result to construct the type we need.
The type of $\mathsf{list_{ts}}$ is
\begin{align*}
\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')}
&= \mathsf{GetType\, B\, t \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'} \\
&= \mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'}
\end{align*}
Thus, we define
\begin{align*}
  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (nothing , lift_{ts'}) = nothing} \\
  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')}
\end{align*}
where $\mathsf{e : Equal\, A\, (List\,B)}$, $\mathsf{T' : LType\, B}$
and $\mathsf{lift_{ts'} : List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'}$.

\subsection{Defining $\mathsf{LTermLKT}$}

{\color{red} Maybe this section can be deleted by assuming $\mathsf{LTerm^{\wedge}\, A\, K_\top\,t = K_\top}$.
Maybe we can say this derives from parametricity.
Currently we say that $\mathsf{K_\top \times K_\top = K_\top}$ based on the fact that products are a built-in type and so this seems to be obviously true.}

The last piece of infrastructure we need in order to define $\mathsf{getTypeProof}$ is a function 
  \mathsf{LTermLKT : \forall\, (A : Set)\, (t : LTerm\, A) \to LTerm^{\wedge}\, A\, K_\top\,t}
that provides a 
proof of $\mathsf{LTerm^{\wedge}\,A\,K_\top\,t}$ for any term $\mathsf{t : LTerm\, A}$. 
Because $\mathsf{LTerm^{\wedge}}$ is defined in terms of $\mathsf{LType^{\wedge}}$, $\mathsf{Arr^{\wedge}}$, 
and $\mathsf{List^{\wedge}}$, we need an analogous function for each of these liftings as well
(respectively, $\mathsf{LTypeLKT}$, $\mathsf{ArrLKT}$ and $\mathsf{ListLKT}$). 
We only give the definition of $\mathsf{LTermLKT}$, but the definitions for 
$\mathsf{LTypeLKT}$, $\mathsf{ArrLKT}$ and $\mathsf{ListLKT}$ are analogous. 
We define $\mathsf{LTermLKT}$ by pattern matching on the lambda term $\mathsf{t}$:
\item For the $\mathsf{var}$ case, let $\mathsf{t = var\, s\, T}$ for $\mathsf{s : String}$ and $\mathsf{T : LType\,A}$,
and define
  \mathsf{LTermLKT\,A\,(var\,s\,T) = LTypeLKT\,A\,T}
\item For the $\mathsf{abs}$ case, let $\mathsf{t = abs \,B \,C \, e \,s \,T \, t'} $
for $\mathsf{e : Equal\,A\,(B \to C)}$, $\mathsf{s : String}$, $\mathsf{T : LType\,B}$ and $\mathsf{t' : LTerm\,C}$,
and recall the definition of $\mathsf{LTerm^{\wedge}}$ for the $\mathsf{abs}$ constructor, 
instantiating the predicate $\mathsf{Q : A \to Set}$ to $\mathsf{K_\top}$: 
% \,a\,Q_a\, (Abs \,b \,c \,e \,s \,T_b \,t_c)}$, :
\mathsf{LTerm^{\wedge}\,A\,K_\top\,
  (abs \,B \,C \,e \,s \,T \,t')} \\ 
\quad\mathsf{
  = \sum\, [Q' : B \to Set]\, [Q'' : C \to Set]\,
      Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q' \, Q'')\, e
      \times \, LType^{\wedge}\, B\, Q'\, T
      \times \, LTerm^{\wedge}\, C\, Q''\, t' }
So, to define the $\mathsf{abs}$ case of $\mathsf{LTermLKT}$, we need a proof of 
  \mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q' \, Q'')\, e}
i.e., that $\mathsf{K_\top}$ is (extensionally) equal to the lifting $\mathsf{Arr^{\wedge} \, B\, C\, Q' \, Q''}$
for some predicates $\mathsf{Q' : B \to Set}$ and $\mathsf{Q'' : C \to Set}$.
The only reasonable choice for $\mathsf{Q'}$ and $\mathsf{Q''}$ 
is to let both be $\mathsf{K_\top}$, which means we need a proof of
  \mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}
Since we are working with proof-relevant predicates
(i.e., functions into $\mathsf{Set}$ rather than functions into $\mathsf{Bool}$), 
the lifting $\mathsf{Arr^{\wedge} \, B\, C\, K_\top \, K_\top}$ of $\mathsf{K_\top}$ to arrow types
is not identical to $\mathsf{K_\top}$ on arrow types, but the predicates are (extensionally) isomorphic. 
We discuss this issue in more detail at the end of the section. 
For now, we assume a proof 
  \mathsf{EqualLArrKT : Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}
and define the $\mathsf{abs}$ case of $\mathsf{LTermLKT}$ as 
\[ \mathsf{LTermLKT\,A\, (abs\, B \,C \, e \,s \,T \, t') = 
    (K_\top , K_\top , EqualLArrKT , LTypeLKT\, B\, T , LTermLKT\, C\, t')
  }
\item For the $\mathsf{app}$ case, let $\mathsf{t = app\, B\, t' \, t''}$
for $\mathsf{t' : LTerm\,(B \to A)}$ and $\mathsf{t'' : LTerm\,B}$,
and, just as we did for the $\mathsf{abs}$ case, 
recall the definition of $\mathsf{LTerm^{\wedge}\, A\, K_\top\, (app\, B\, t'\, t'')}$
with all of the predicates instantiated with $\mathsf{K_\top}$: 
\mathsf{
      LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t'
      \times LTerm^{\wedge}\, B\, K_\top\, t''}
The second component can be given using $\mathsf{LTermLKT}$, and we can 
define the first component using a proof of
\mathsf{LTerm^{\wedge}\, (B \to A)\, K_\top\, t'}
and a map-like function 
  \mathsf{LTermLEqualMap : \forall\, (A : Set)\, (Q\,Q' : A \to Set) 

>>>>

{\bf Related Work\/} Various techniques for deriving induction rules
for data types that go beyond ADTs have been studied. For example, Fu
and Selinger~\cite{fs18} show, via examples, how to derive induction
rules for arbitrary nested types. Unfortunately, however, their
technique is rather {\em ad hoc}, so is unclear how to generalize it
to nested types other than the specific ones in the
examples. Moreover, it actually derives induction rules for data types
{\em related} to the original nested types rather than for the
original nested types themselves, and it is unclear whether or not the
derived rules are sufficiently expressive to prove all results about
the original nested types that we would expect to be provable by
induction. This latter point echoes the issue with Coq-derived
induction rule for rose trees raised in Section~\ref{sec:intro}, which
has the unfortunate effect of forcing users to manually write
induction (and other) rules for such types for use in that
system. Tassi~\cite{tas19} has done exactly that, deriving induction
rules for data type definitions in Coq using unary
parametricity. Tassi's technique seems to be essentially equivalent to
that of~\cite{jp19} for nested types, although he does not permit true
nesting. More recently, Ullrich~\cite{ull20} has implemented a plugin
in MetaCoq to generate induction rules for nested types. This plugin
is also based on unary parametricity and, again, true nesting is not
permitted.  As far as we know, no attempt has yet been made to extend
either implementation to GADTs. In fact, we know of no work other than
that reported here that specifically addresses induction rules for
(deep) GADTs.

@@

Although a truly nested type admits only a single induction rule, it
is worth noting that for those nested types that do admit distinct
structural induction and deep induction rules, the latter generalizes
the former. Indeed, the structural induction rule for such a nested
type is recoverable from its deep induction rule by taking the custom
predicates on its data of primitive types to be constantly
$\mathsf{True}$-valued predicates. This instantiation ensures that the
resulting induction rule only inspects the top-level structure of its
argument, rather than the contents of that structure, which exactly
coincides with what structural induction should do.
%A concrete example of such a derivation is given in
%Section~\ref{sec:ind-equal}.

@@

The equality constraints engendered by GADTs' data constructors makes
deriving (deep) induction rules for then more involved than for ADTs
and other nested types. Nevertheless, we show in this section how to
do so. We first illustrate the key components of our approach by
deriving deep induction rules for the three specific GADTs introduced
in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
abstract these to a general framework that can be applied to any deep
GADT that is not truly nested. As hinted above, the predicate lifting
for the equality GADT plays a central role in deriving both structural
and deep induction rules for more general GADTs.

@@

To prove that this rule is sound we provide a witness
$\mathsf{dIndSeq}$ inhabiting the type in~\eqref{eq:ind-seq} as
follows:
\[\begin{array}{lll}
\mathsf{dIndSeq\;P\;cconst\;cpair\;A\;Q_A\;(const\,a)\;liftA}  &
    = &  \mathsf{cconst\;A\;Q_A\;a\;liftA}\\
\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,
    (Q_B,Q_C, liftE, liftB, liftC)}  & = &
\mathsf{cpair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}
\end{array}\]
In the first clause above, $\mathsf{a : A}$, $\mathsf{Q_A : A \to
  Set}$, $\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) =
  Q_A\,a}$. In the second, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C :
  C \to Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, $\mathsf{liftE :
  Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e}$,
$\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$, and $\mathsf{liftC :}$
$\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together ensure that
$\mathsf{(Q_B, Q_C, liftE, liftB, liftC) :
  Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$ --- and
$\mathsf{p_B} =\mathsf{dIndSeq\,P\,cconst\,cpair\,B\,Q_B\,s_B\,liftB :
  P\,B\,Q_B\,s_B}$ and $\mathsf{p_C}
=\mathsf{dIndSeq\,P\,cconst\,cpair\,C\,Q_C\,s_C\,liftC :
  P\,C\,Q_C\,s_C}$.

@@

In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
derived deep induction rules for GADTs that are not truly
nested. Since both nested types and GADTs without true nesting admit
deep induction rules, we might expect that truly nested GADTs would as
well. Surprisingly, however, they do not. That is, our results from
the previous section are the strongest possible. Indeed, the induction
rule for a data type generally relies on (unary) parametricity of the
model interpreting it, and deep induction for a truly nested type or a
truly nested GADT also relies on this interpretation being functorial.
But, whereas nested types can admit a functorial parametric semantics,
GADTs cannot admit both functorial and parametric semantics at the
same time~\cite{jgj21}. In this section we show how deep induction for
truly nested GADTs nesting goes wrong by analyzing the following
concrete example of such a type.\looseness=-1

@@

In this section we use deep induction for the $\mathsf{LTerm}$ GADT
from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  For
this we use the following predicate, which takes a lambda term and
produces the type of its possible types:
\begin{align*}
  &\mathsf{GetType : \forall \, (A : Set) \to LTerm\,A \to Set} \\
  &\mathsf{GetType \,A \,t = Maybe \, (LType \, A)}
\end{align*}
The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
data type to represent potential failure. It is defined by:

\begin{equation}\label{eq:maybe}
\begin{array}{l}
\mathsf{data\ Maybe : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;nothing :\, Maybe\,A}\\
\mathsf{\;\;\;\;\;\;\;\;\;\;\;\;just :\, A \to Maybe\,A}
\end{array}
\end{equation}

We want to show that $\mathsf{GetType}$ is satisfied by every element
of $\mathsf{LTerm\,A}$, i.e., we want to prove:
\[ \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to
   GetType \,A \,t}\] This cannot be achieved without deep induction,
which is needed to apply the induction hypothesis to the individual
terms in the list of terms that the data constructor $\mathsf{list}$
takes as an argument. Using the deep induction rule
$\mathsf{dIndLTerm}$ from Section~\ref{sec:ind-lam} we define
$\mathsf{getTypeProof}$ by
\[ \mathsf{getTypeProof \,A \,t = 
    dIndLTerm\, P \, cvar\, cabs\, capp\, clistc\, A\, K_\top\, t\,
    (LTerm^\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
$\mathsf{P}$ is the polymorphic predicate $\mathsf{\lambda \, (A:
  Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
  A)}$, $\mathsf{K_\top}$ is the constantly {\color{red}
  $\mathsf{\top}$}-valued predicate, and $\mathsf{LTerm^\wedge KT :
  \forall\, (A : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\,
  K_\top\,t}$ is a term, to be defined below, witnessing that
$\mathsf{K_\top}$ can be lifted to all terms. We also need the
applications to $\mathsf{P}$ of each of the induction hypotheses from
Section~\ref{sec:ind-lam}. These are:
\[\begin{array}{lll}
\mathsf{cvar} & : & \mathsf{\forall\, (A : Set)\, (Q_A : A \to Set)\,
  (s : String)\, (T_A : LType\, A) \to LType^{\wedge} \, A\, Q_A\,
  T_A\, \to Maybe\, (LType\, A)}\\
 \mathsf{cabs} & : & \mathsf{\forall\, (A \, B\, C: Set)\, (Q_A : A
   \to Set)\, (Q_B : B \to Set)\, (Q_C : C \to Set)} \\ 
& &    \quad\mathsf{(e : Equal\, A\, (B \to C))\, (s : String)\, (T_B
   : LType\, B)\, (t_C : LTerm \, C)} \\  
& & \quad\quad\mathsf{\to Equal^{\wedge}\, A\, (B \to C)\, Q_A\,
   (Arr^{\wedge}\, B\, C\, Q_B\, Q_C)\, e  \to LType^{\wedge} \, B\,
   Q_B\, T_B  \to Maybe\, (LType\,C) \to Maybe\, (LType\, A)}\\
\mathsf{capp} & : & \mathsf{\forall\, (A \, B : Set)\, (Q_A : A \to
  Set)\, (Q_B : B \to Set)\,  (t_{BA}: LTerm\, (B \to A))\, (t_B :
  LTerm\, B)} \\  
& & \quad\mathsf{\to Maybe\, (LType\, (B \to A))\to Maybe\, (LType\,
      B) \to Maybe\, (LType\, A)}\\
\mathsf{clist} & : & \mathsf{\forall\, (A \, B : Set)\, (Q_A : A \to
  Set)\, (Q_B : B \to Set)\, (e : Equal\, A\, (List\, B))\, (ts : 
  List\, (LTerm\, B))}\\  
& & \quad\mathsf{\to Equal^{\wedge}\, A\, (List\,B) \, Q_A\,
  (List^{\wedge}\, B\, Q_B)\, e \to List^{\wedge}\, (LTerm\, B)\,
  (GetType\, B)\, ts \to Maybe\, (LType\, A)}
In the first clause, $\mathsf{cvar}$ returns $\mathsf{just\,T_A}$.  In
the second clause, $\mathsf{cabs}$ returns $\mathsf{nothing}$ if its
final argument is $\mathsf{nothing}$ and $\mathsf{cabs\, A \,B \,C
  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, Eq^\wedge\,
%  LLTy^\wedge B\,
L\, (just \,T_C) = just \, (arr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
In the third clause, $\mathsf{capp\, A \,B \, Q_A \, Q_B \, t_{BA}\,
  t_A \, (just \, (arr\, .B \,.A\, rfl\, T_B\, T_A))\, mb = just\,
  T_A}$ and $\mathsf{capp}$ returns $\mathsf{nothing}$ otherwise.  In
the fourth clause, we must use $\mathsf{List^{\wedge}\, (LTerm\, B)\,
  (GetType\, B)\, ts}$ to extract the type of the head of
$\mathsf{ts}$ (from which we can deduce the type of the list). When
$\mathsf{ts} = \mathsf{nil}$ we define $\mathsf{clist\, A\, B\, Q\,
  Q'\, e\, nil \, liftE\, lift_{ts} = nothing}$, where $\mathsf{liftE
  : Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\,
  e}$, and $\mathsf{lift_{ts} : List^{\wedge}\, (LTerm\, B)\,
  (GetType\, B)\, ts}$.  When $\mathsf{ts} = \mathsf{cons\,t\,ts'}$
% The interesting case is
%  $\mathsf{clist}$, in which we have to use the results of
%  $\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$ in order
%  to extract the type of one of the terms in the list.  To define
%  $\mathsf{clist}$ we pattern-match on the list of terms
%  $\mathsf{ts}$.  If $\mathsf{ts}$ is the empty list $\mathsf{nil}$,
%  we cannot extract a type, so we return $\mathsf{nothing}$:
%\[
%  \mathsf{clist\, A\, B\, Q\, Q'\, e\, nil \, liftE\, lift_{ts} = nothing}
%\]
%where $\mathsf{liftE : Equal^{\wedge}\, A\, (List\,B) \, Q\,
%  (List^{\wedge}\, B\, Q')\, e}$ and $\mathsf{lift_{ts} :
%  List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.
%If $\mathsf{ts}$ is a non-empty list $\mathsf{cons\,t\,ts'}$, we
%pattern match on
%$\mathsf{list_{ts}}$ and use the result to construct
%the type we need.  The type of $\mathsf{list_{ts}}$ is
the type of $\mathsf{list_{ts}}$ becomes $\mathsf{List^{\wedge}\,
  (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')} = \mathsf{GetType\,
  B\, t \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'} =
\mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\,
  (GetType\, B)\, ts'}$. We pattern match on the first component
of the pair to define\looseness=-1
%\[\begin{array}{lll}
%\mathsf{list_{ts}} & : &  
%  \mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')}\\
%& = & \mathsf{GetType\, B\, t \times List^{\wedge}\, (LTerm\, B)\,
%    (GetType\, B)\, ts'}\\ 
%&= & \mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\,
%    (GetType\, B)\, ts'} 
%\end{array}\]
\begin{align*}
  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
    (nothing , lift_{ts'}) = nothing} \\ 
  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
    (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')} 
\end{align*}
Here $\mathsf{e : Equal\, A\, (List\,B)}$, $\mathsf{T' : LType\, B}$,
and $\mathsf{lift_{ts'} : List^{\wedge}\, (LTerm\, B)\, (GetType\,
  B)\, ts'}$.

To finish defining $\mathsf{getTypeProof}$ we still need a proof

\[ \mathsf{LTerm^\wedge KT : \forall\, (A : Set)\, (t : LTerm\, A) \to
  LTerm^{\wedge}\, A\, K_\top\,t} \]

\noindent
Since $\mathsf{LTerm^\wedge}$ is
defined in terms of $\mathsf{LType^\wedge}$ and $\mathsf{Arr^\wedge}$,
and since $\mathsf{LType^\wedge}$ is also defined in terms of
$\mathsf{List^\wedge}$, we need analogous functions
$\mathsf{LType^\wedge KT}$, $\mathsf{Arr^\wedge KT}$ and
$\mathsf{List^\wedge KT}$, respectively, for each of these liftings as
well.  We only give the definition of $\mathsf{LTerm^\wedge KT}$ here
since $\mathsf{LType^\wedge KT}$, $\mathsf{Arr^\wedge KT}$, and
$\mathsf{List^\wedge KT}$ are defined analogously. We have:
\begin{itemize}
\item If $\mathsf{s : String}$ and $\mathsf{T : LType\,A}$ we define
  $\mathsf{LTerm^\wedge KT\,A\,(var\,s\,T)} = \mathsf{LType^\wedge
  KT\,A\,T}$.
\item If $\mathsf{e : Equal\,A\,(B \to C)}$, $\mathsf{s : String}$,
  $\mathsf{T : LType\,B}$, and $\mathsf{t' : LTerm\,C}$ we need to
  define $\mathsf{LTerm^\wedge KT\,A\, (abs\, B \,C \, e \,s \,T \,
    t')}$ of type

\vspace*{-0.15in}
  \mathsf{LTerm^{\wedge}\,A\,K_\top\, (abs \,B \,C \,e \,s \,T
  \,t')}\\
\quad \mathsf{ = \exists [Q_B]\, [Q_C]\, Equal^{\wedge} \, A\, (B \to
    C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e \times \,
    LType^{\wedge}\, B\, Q_B\, T \times \, LTerm^{\wedge}\, C\, Q_C\,
    t' }
where $\mathsf{K_\top : A \to Set}$, $\mathsf{Q_B : B \to Set}$, and
$\mathsf{Q_C : C \to Set}$.  The only reasonable choice is to let both
$\mathsf{Q_B}$ and $\mathsf{Q_C}$ be $\mathsf{K_\top}$, which means we
need proofs of $\mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\,
  (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}$,
$\mathsf{LType^{\wedge}\, B\, K_\top\, T}$ and
$\mathsf{LTerm^{\wedge}\, C\, K_\top\, t'}$.  We take
$\mathsf{LType^\wedge KT\, B\, T}$ and $\mathsf{LTerm^\wedge KT\, C\,
  t')}$ for the latter two proofs. For the former we note that, since
we are working with proof-relevant predicates, the lifting
$\mathsf{Arr^{\wedge} \, B\, C\, K_\top \, K_\top}$ of
$\mathsf{K_\top}$ to arrow types is not identical to $\mathsf{K_\top}$
on arrow types but rather (extensionally) isomorphic.  We discuss this
issue in more detail at the end of the section, but for now we simply
assume a proof $\mathsf{Equal^\wedge ArrKT : Equal^{\wedge} \, A\, (B
  \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}$
